The Two Dualities of Computation: Negative and Fractional Types

### 2012-03-08 rational.tex
semantics of pi

%subcode-inline{bnf-inline} regex \{\{(((\}[^\}])|[^\}])*)\}\} name main include langRev
%! Gx = \Gamma^{\times}
%! G = \Gamma
%! [] = \Box
%! |-->* = \mapsto^{*}
%! |-->> = \mapsto_{\ggg}
%! |--> = \mapsto
%! |- = \vdash
%! <><> = \approx
%! ==> = \Longrightarrow
%! <== = \Longleftarrow
%! <=> = \Longleftrightarrow
%! <-> = \leftrightarrow
%! ~> = \leadsto
%! -o = \multimap
%! ::= = &::=&
%! /= = \neq
%! @@ = \mu
%! forall = \forall
%! exists = \exists
%! empty = \epsilon
%! langRev = \Pi
%! langRevT = \Pi^{o}
%! langRevEE = \Pi^{\eta\epsilon}
%! theseus = Theseus
%! sqrt(x) = \sqrt{#x}
%! surd(p,x) = \sqrt[#p]{#x}
%! inv(x) = \frac{1}{#x}
%! frac(x,y) = \frac{#x}{#y}
%! * = \times

@@

For the non-primitive combinators, we express the semantics using evaluation
contexts which are used to represent the ``rest of the computation.'' More
precisely, to calculate the result of applying combinator $c$ to value $v$,
we start a small-step abstract machine in the configuration {{<c,v,[]>}}. The
transitions of the machine (see below) are unusual in the sense that they
always maintain full information about {{c}}, although this information may
be split in the first and last component of the machine. The {{v}} component
of the machine is the one that captures the current value as it is flowing in
the combinator {{c}}. When  scattered in the machine will examine {{c}} and make
transitions based on the structure of {{c}}. When the evaluation of {{c}} is
complete, the configuration should be {{[c,v,[]]}} where the {{c}} is
restored.

This is a mess. Perhaps give a small example and explain things based on the
example before giving the full transitions.

@@

%subcode{opsem} include main
%! columnStyle = rclr
% <iso, v, C> &|-->& {[iso, iso(v), C]}
% <c1(;)c2, v, C> &|-->& <c1, v, Fst C c2>
% {[c1, v, Fst C c2]} &|-->& <c2, v, Snd c1 C>
% {[c2, v, Snd c1 C]} &|-->& {[ c1(;)c2, v, C ]}
% <c1(+)c2, left v, C> &|-->& <c1, v, LeftP C c2>
% {[ c1, v, LeftP C c2 ]} &|-->& {[c1 (+) c2, left v, C ]}
% <c1(+)c2, right v, C> &|-->& <c2, v, RightP c1 C>
% {[ c2, v, RightP c1 C ]} &|-->& {[c1 (+) c2, right v, C ]}
% <c1(*)c2, (v1, v2), C> &|-->& <c1, v1, LeftT C c2 v2>
% {[ c1, v1, LeftT C c2 v2 ]} &|-->& <c2, v2, RightT c1 v1 C>
% {[ c2, v2, RightT c1 v1 C ]} &|-->& {[ c1 (*) c2, (v1, v2), C ]}

>>>>

%subcode-inline{bnf-inline} regex \{\{(((\}[^\}])|[^\}])*)\}\} name main include langRev
%! Gx = \Gamma^{\times}
%! G = \Gamma
%! [] = \Box
%! |-->* = \mapsto^{*}
%! |-->> = \mapsto_{\ggg}
%! |--> = \mapsto
%! <--| = \mapsfrom
%! |- = \vdash
%! <><> = \approx
%! ==> = \Longrightarrow
%! <== = \Longleftarrow
%! <=> = \Longleftrightarrow
%! <-> = \leftrightarrow
%! ~> = \leadsto
%! -o = \multimap
%! ::= = &::=&
%! /= = \neq
%! @@ = \mu
%! forall = \forall
%! exists = \exists
%! empty = \epsilon
%! langRev = \Pi
%! langRevT = \Pi^{o}
%! langRevEE = \Pi^{\eta\epsilon}
%! theseus = Theseus
%! sqrt(x) = \sqrt{#x}
%! surd(p,x) = \sqrt[#p]{#x}
%! inv(x) = \frac{1}{#x}
%! frac(x,y) = \frac{#x}{#y}
%! * = \times

@@

The reductions for the primitive isomorphisms are exactly the same as
have been persented before \cite{infeffects}. The reductions for the
closure combinators are however presented in a small-setp operational
style. The following are the definition for evaluation contexts and
machine states:

@@

The reduction rules below essentially track the flow of particles
through their circuit. The start machine state, {{<c,v,[]>}}, denotes
the particle/value {{v}} about to be evluated by the circuit
{{c}}. The end machine state, {{[c, v, [] ]}}, denotes the situation
where the particle {{v}} has exited the circuit {{c}}.

%% For the non-primitive combinators, we express the semantics using evaluation
%% contexts which are used to represent the ``rest of the computation.'' More
%% precisely, to calculate the result of applying combinator $c$ to value $v$,
%% we start a small-step abstract machine in the configuration {{<c,v,[]>}}. The
%% transitions of the machine (see below) are unusual in the sense that they
%% always maintain full information about {{c}}, although this information may
%% be split in the first and last component of the machine. The {{v}} component
%% of the machine is the one that captures the current value as it is flowing in
%% the combinator {{c}}. When  scattered in the machine will examine {{c}} and make
%% transitions based on the structure of {{c}}. When the evaluation of {{c}} is
%% complete, the configuration should be {{[c,v,[]]}} where the {{c}} is
%% restored.

%% This is a mess. Perhaps give a small example and explain things based on the
%% example before giving the full transitions.

%subcode{opsem} include main
%! columnStyle = rclr
% <iso, v, C> &|-->& {[iso, v', C]} & (1)
% & & where iso v |--> v' & (2) 
% <c1(;)c2, v, C> &|-->& <c1, v, Fst C c2> & (3) 
% {[c1, v, Fst C c2]} &|-->& <c2, v, Snd c1 C> & (4) 
% {[c2, v, Snd c1 C]} &|-->& {[ c1(;)c2, v, C ]} & (5) 
% <c1(+)c2, left v, C> &|-->& <c1, v, LeftP C c2> & (6) 
% {[ c1, v, LeftP C c2 ]} &|-->& {[c1 (+) c2, left v, C ]} & (7)
% <c1(+)c2, right v, C> &|-->& <c2, v, RightP c1 C> & (8)
% {[ c2, v, RightP c1 C ]} &|-->& {[c1 (+) c2, right v, C ]} & (9)
% <c1(*)c2, (v1, v2), C> &|-->& <c1, v1, LeftT C c2 v2> & (10)
% {[ c1, v1, LeftT C c2 v2 ]} &|-->& <c2, v2, RightT c1 v1 C> & (11) 
% {[ c2, v2, RightT c1 v1 C ]} &|-->& {[ c1 (*) c2, (v1, v2), C ]} & (12)

The rule (1) describes evaluation by a primitive isomorphism. Rules
(2), (3) and (4) deal with sequential evaluation. Rule (2) says that
for the value to flow through the sequence {{c1 (;) c2}} we should
first flow through {{c1}} with {{c2}} pending in the context and rule
(3) says the value that exits from {{c1}} should proceed to flow
through {{c2}}. Rule (4) says that if the value {{v}} has exited
{{c2}} as well, then it is equivalent to having exited from
{{c1(;)c2}}.

Rules (5) to (9) deal with {{c1 (+) c2}} in the same way. In the case
of sums, the shape of the value {{left v}}/{{right v}} determines
whether path {{c1}} or path {{c2}} is taken. Rules (10), (11) and (12)
deal with {{c1 (*) c2}} similarly. In the case of products the value
should have the form {{(v1, v2)}} where {{v1}} flows through {{c1}}
and {{v2}} flows through {{c2}}. Both these paths are entirely
independant of each other and we could chosen to evaluate either
first, or evaluate both in parallel. In this presentation we have
chosen to follow {{c1}] first, but this choice is entirely artificial.

Finally, we end by noting the abstract machine here is interesting is
that it is a reversible machine. We could change reductions `{{|-->}}'
to run backwards `{{<--|}}', thereby computing the start state from the
stop state. We will take advantange of this property in
Sec. \ref{sec:rat}.
